---
title: "`r params$clone` Bactdating test run gubbins (`r params$iterations` iterations)"
params:
    clone: SL_1
    iterations: 100000
    seed: 3
    tree_for_bactdate: rooted_tree_rec
    n_alignment_sites: 1
    best_model: relaxedgamma
    run_mode: full
    na_year: include
---

```{r setup, include=FALSE, message=F}
knitr::opts_chunk$set(echo = TRUE)
```

## Table of contents
- [Root-to-tip regression](#roottotip)  
- [BactDating MCMC model selection](#bactdatingmodelselection)
- [BactDating MCMC analysis plots](#bactdatingmcmcplots)


## Set up

```{r set_vars}
clone <- params$clone
clone
seed <- params$seed
seed
chain_length <- params$iterations
chain_length
tree_for_bactdate <- params$tree_for_bactdate
tree_for_bactdate

if (tree_for_bactdate == 'rooted_tree_rec') {
    bd_use_rec = T
} else {
    bd_use_rec = F
}

# Set R project directory
project_dir <- "~/erkison/klebs_dataset/"

# Set genome analyses directory (for gubbins and other output)
genome_analyses_dir <- file.path(project_dir, clone)

# Set seed
set.seed(seed)

# Set burnin proportion for MCMC runs
burnin_prop <- 0.5
run_mode <- params$run_mode
na_year <- params$na_year

```


## Load packages & track versions

<details>
<summary>Click to expand</summary>
    ```{r load_packages}
    # R version
    R.Version()$version.string
    
    # Reporting
    library(knitr)
    packageVersion("knitr")
    
    # Wrangling
    library(tidyverse)
    packageVersion("tidyverse")
    
    library(lubridate)
    packageVersion("lubridate")
    
    library(coda)
    packageVersion("coda")
    
    # Phylo
    library(BactDating)
    packageVersion("BactDating")
    
    library(ape)
    packageVersion("ape")
    
    library(ggtree)
    packageVersion("ggtree")
    
    library(treeio)
    packageVersion("treeio")
    
    # Custom functions
    source(file.path(project_dir, "bactdating/scripts/bactdating_functions.R"))
    ```
</details>  

<br>

## Import data

<details>
<summary>Click to expand</summary>
```{r import_data}
# Read in all clones data
clone_data <-
    readr::read_tsv(file.path(project_dir, clone, paste0(clone, "_data.tsv")),
                    show_col_types = F)

# get_clone_data_for_bactdate

# Get and format isolate dates and gubbins tree
assign(glue::glue("{clone}_data_for_bactdate"),
       prepare_bactdate_data(clone_data, 
                             gubbins_dir=file.path(genome_analyses_dir, "gubbins"))
)

tree <- get(glue::glue("{clone}_data_for_bactdate"))$tree
dates_rtt <- get(glue::glue("{clone}_data_for_bactdate"))$dates_rtt
dates_bactdate <- get(glue::glue("{clone}_data_for_bactdate"))$dates_bactdate

if (na_year == "exclude"){
    # Exclude leaves with missing year info (NA)
    index_dates_to_drop <- which(is.na(dates_rtt))
    if(length(index_dates_to_drop) >= 1){
        dates_rtt <- dates_rtt[-index_dates_to_drop]
        dates_bactdate <- dates_bactdate[-index_dates_to_drop,]
        tree <- treeio::drop.tip(tree, names(index_dates_to_drop)) %>% 
            phangorn::midpoint()
        print(glue::glue("Excluded {length(index_dates_to_drop)} samples with NA year."))
    }
}

# optionally exclude other samples, if specified
other_ids_to_drop <- c("")
idx_other_ids_to_drop <- which(names(dates_rtt) %in% other_ids_to_drop)
if (length(idx_other_ids_to_drop) >= 1) {
    dates_rtt <- dates_rtt[-idx_other_ids_to_drop]
    dates_bactdate <- dates_bactdate[-idx_other_ids_to_drop,]
    tree <- treeio::drop.tip(tree, other_ids_to_drop) %>% 
            phangorn::midpoint()
    print(glue::glue("Excluded {paste0(other_ids_to_drop, collapse=', ')}."))
}

# Plot tree
plot(tree, show.tip.label = F)
ape::axisPhylo(backward=F)

# initial tree rooting based on best root-to-tip correlation
rooted_tree <- BactDating::initRoot(tree, dates_rtt)
rooted_tree_rec <- BactDating::initRoot(tree, dates_rtt, useRec = T)

```
</details>  

<br>

## Root-to-tip regression analysis for temporal signal {#roottotip}

<details>
<summary>Click to expand</summary>
```{r root_to_tip_regression}
# Root to tip regression (unrooted init tree)
BactDating::roottotip(tree, dates_rtt) 
# Root to tip regression (rooted init tree with )  
BactDating::roottotip(rooted_tree, dates_rtt)
# Root to tip regression (rooted init tree with gubbins recomb info)
BactDating::roottotip(rooted_tree_rec, dates_rtt)

# set what tree to use for BactDating runs
tree_for_bactdate <- get(tree_for_bactdate) # use preset params
# or simply reset based on RTT results. E.g., tree_for_bactdate <- rooted_tree_rec
tree_for_bactdate
```
</details>

<br>

## MCMC BactDating model selection {#bactdatingmodelselection}

<details>
<summary>Click to expand</summary>
```{r strict}

if(! ape::is.rooted(tree_for_bactdate)){
    tree_for_bactdate <- phangorn::midpoint(tree_for_bactdate)
}

# Strict clock
strict_out <- file.path(project_dir, "bactdating", clone, "with_gubbins",
                        glue::glue("{clone}_strict_mcmc_{chain_length}.rds"))
if (run_mode == "resume" & file.exists(strict_out)){
    bactdating_strict_mcmc <- readRDS(strict_out)
} else {
    # Run bactdate
    bactdating_strict_mcmc <- BactDating::bactdate(
        tree_for_bactdate, dates_bactdate, model='mixedgamma', nbIts=chain_length,
        useRec=bd_use_rec, updateRoot=T, showProgress=F)
}
bactdating_strict_mcmc
plot(bactdating_strict_mcmc, type="trace")
try(plot(bactdating_strict_mcmc, type="scatter"))
bactdating_strict_mcmc$pstrict
saveRDS(bactdating_strict_mcmc, file=strict_out)

bd_strict_mcmc_summary <- extract_bactdating_summary(bactdating_strict_mcmc, 
                                                     clone, burnin_prop)
write_tsv(bd_strict_mcmc_summary,
          file.path(sub(".rds", ".tsv", strict_out)))
```

```{r relaxed}
# Relaxed clock
relaxed_out <- file.path(project_dir, "bactdating", clone, "with_gubbins",
                         glue::glue("{clone}_relaxed_mcmc_{chain_length}.rds"))
if (run_mode == "resume" & file.exists(relaxed_out)){
    bactdating_relaxed_mcmc <- readRDS(relaxed_out)
} else {
    # Run bactdate
    bactdating_relaxed_mcmc <- BactDating::bactdate(
        tree_for_bactdate, dates_bactdate, model='relaxedgamma', nbIts=chain_length,
        useRec=bd_use_rec, updateRoot=T, showProgress=F)
}
bactdating_relaxed_mcmc
plot(bactdating_relaxed_mcmc,type='trace')
try(plot(bactdating_relaxed_mcmc, type="scatter"))
bactdating_relaxed_mcmc$prelaxed
saveRDS(bactdating_relaxed_mcmc, file=relaxed_out)

bd_relaxed_mcmc_summary <- extract_bactdating_summary(bactdating_relaxed_mcmc, 
                                                     clone, burnin_prop)
write_tsv(bd_relaxed_mcmc_summary,
          file.path(sub(".rds", ".tsv", relaxed_out)))
```

```{r model_selection}
# Check if ESS sufficient for model comparison
coda::effectiveSize(BactDating::as.mcmc.resBactDating(bactdating_strict_mcmc, burnin = burnin_prop))
coda::effectiveSize(BactDating::as.mcmc.resBactDating(bactdating_relaxed_mcmc, burnin = burnin_prop))

# Compare models
BactDating::modelcompare(bactdating_strict_mcmc,bactdating_relaxed_mcmc)

```
</details>

<br>

## MCMC plots both models {#bactdatingmcmcplots}

<details>
<summary>Click to expand</summary>
```{r mcmc_plots}

# Plot strict clock model tree with CI
plot(bactdating_strict_mcmc,'treeCI', show.tip.label = F)

# Plot strict clock most likely root
plot(bactdating_strict_mcmc,'treeRoot', show.tip.label=F)


# Plot relaxed clock tree with CI
plot(bactdating_relaxed_mcmc,'treeCI', show.tip.label = F)

# Plot relaxed clock most likely root
plot(bactdating_relaxed_mcmc,'treeRoot', show.tip.label=F)

```
</details>

<br>

